---
title: "Dynamic spatio-temporal models for post-fire big sagebrush recovery"
author: "Andrii Zaiats, Rongsong Liu, David Pilliod, Megan Cattau, Trevor Caughlin"
date: "4/11/2020"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
---
*** 

Mechanistic models for spatio-temporal ecological processes  
PDE based dynamics  
$$ \frac{\partial u}{\partial t} = \mathcal{H}(u, \omega, \theta)$$
which in two dimentions
$$\mathcal{H}(u,\omega,\theta) \equiv \bigg(\frac{\partial^2}{\partial x^2}+\frac{\partial^2}{\partial y^2}\bigg)\delta (x,y) u(x,y)$$
where $\delta (x,y)$ is a spatially varying diffusion parameter.  

***

Using forward difference in time and central difference in space we get the following integro-difference equation (Haberman 1987):

$$
\begin{aligned}
u_{t,(x,y)} = & u_{t-\Delta t,(x,y)} \bigg[1 - 2\delta_{(x,y)} \bigg(\frac{\Delta t}{\Delta x}+\frac{\Delta t}{\Delta y}\bigg)\bigg] \\ 
  & +u_{t-\Delta t,(x-\Delta x,y)} \bigg[\frac{\Delta t}{\Delta x} \{\delta_{(x,y)}-\bigg(\frac{\delta_{(x+\Delta x,y)}-\delta_{(x-\Delta x,y)}}{4}\bigg) \} \bigg] \\
  & +u_{t-\Delta t,(x+\Delta x,y)} \bigg[\frac{\Delta t}{\Delta x} \{\delta_{(x,y)}+\bigg(\frac{\delta_{(x+\Delta x,y)}-\delta_{(x-\Delta x,y)}}{4}\bigg)\}\bigg] \\
  & +u_{t-\Delta t,(x,y+\Delta y)} \bigg[\frac{\Delta t}{\Delta y^2} \{\delta_{(x,y)}+\bigg(\frac{\delta_{(x,y+\Delta y)}-\delta_{(x,y-\Delta y)}}{4}\bigg) \} \bigg] \\
  & +u_{t-\Delta t,(x,y-\Delta y)} \bigg[\frac{\Delta t}{\Delta y^2} \{\delta_{(x,y)}-\bigg(\frac{\delta_{(x,y+\Delta y)}-\delta_{(x,y-\Delta y)}}{4}\bigg) \} \bigg] \\
    & + \eta_{t,(x,y)}
\end{aligned}
$$

Which can be expressed as:
$$ u_t = \mathbf{H}(\delta,\Delta t, \Delta x, \Delta y)u_t+\eta_t $$

***  

To adapt the model to capture population growth, we can add $\mathcal{f}(u)$ resulting in a spatial logistic growth model:
$$ u_t = \mathbf{H}(\delta,\Delta t, \Delta x, \Delta y)u_t+\gamma_0u_{t-1}+\frac{\gamma_0}{\gamma_1}u^2_{t-1}+\eta_t $$
```{r, echo=FALSE, include=TRUE, message=FALSE}
library(raster)
library(rgdal)
makeH <- function(NN,Dvec,dx,dy){
  H <- matrix(0,dim(NN)[1],dim(NN)[1])
  for(i in 1:dim(H)[1]){
    if(sum(NN[i,]!=0) == 4){
      H[i,i] <- 1 - 2*Dvec[i]*(1/dx^2 + 1/dy^2)
      if(NN[i,1]>0){H[i,NN[i,1]] <- (Dvec[i] - (Dvec[NN[i,2]] - Dvec[NN[i,1]])/4)/dx^2}
      if(NN[i,2]>0){H[i,NN[i,2]] <- (Dvec[i] + (Dvec[NN[i,2]] - Dvec[NN[i,1]])/4)/dx^2}
      if(NN[i,3]>0){H[i,NN[i,3]] <- (Dvec[i] + (Dvec[NN[i,4]] - Dvec[NN[i,3]])/4)/dy^2}
      if(NN[i,4]>0){H[i,NN[i,4]] <- (Dvec[i] - (Dvec[NN[i,4]] - Dvec[NN[i,3]])/4)/dy^2}}
  }
  H
}
wrap = function(x,poly){
  row=dim(poly)[1]
  col=dim(poly)[2]
  mat=matrix(x, row,col, byrow=TRUE)
  mat
}
neighborhood <- function(raster){
  nn <- matrix(NA,length(raster[]),4)
  for(i in 1:dim(nn)[1]){
    loc <- adjacent(raster,i)[,2]
    ln <- loc[which((loc+1)==i)]
    rn <- loc[which((loc-1)==i)]
    bn <- loc[which((loc-dim(raster)[2])==i)]
    tn <- loc[which((loc+dim(raster)[2])==i)]
    nn[i,1] <- ifelse(length(ln)>0,ln,0)
    nn[i,2] <- ifelse(length(rn)>0,rn,0)
    nn[i,3] <- ifelse(length(tn)>0,tn,0)
    nn[i,4] <- ifelse(length(bn)>0,bn,0)
  }
  nn
}
# --- load data and transform to N x T formal
years=c(1985:2017)
path="~/MEGA/landsat_backintime/"
file="BogusCreek_1986_clipped_30_30.tif" #1986
r=stack(paste0(path,file))
poly=subset(r,2:14)
values(poly)[values(poly)==101]=NA
df1=as.data.frame(poly,xy=TRUE,na.rm=TRUE)
names(df1)=c('x','y',paste0('year',years[2:14]))

NN=neighborhood(poly[[1]])
ymat=as.matrix(df1[,-c(1,2)])
n=nrow(ymat)
T=ncol(ymat)
dx=30
dy=30
y0=ymat[,1]
r=.8
k=40
```
Deterministic simultation of non-spatial growth: $u_t=\gamma_0u_{t-1}+\frac{\gamma_0}{\gamma_1}u^2_{t-1}$  
```{r, echo=FALSE,include=TRUE,fig.height = 4, fig.width = 6, fig.align = "center"}
ysim=matrix(NA,n,T)
ysim[,1] = y0
for(t in 2:T){
  ysim[,t] = ysim[,t-1] + r*ysim[,t-1]*(1-ysim[,t-1]/k) #
}
matplot(t(ysim),type="l",ylab="Cover, %", xlab="Time, years",)
```

***  

By adding spatial term to the equation:$u_t=\mathbf{H}(\delta,\Delta t, \Delta x, \Delta y)u_t+\gamma_0u_{t-1}+\frac{\gamma_0}{\gamma_1}u^2_{t-1}$. In this case, the recovery in each pixel depends on it's intrinsic growth and neighboring individuals.   

```{r, echo=FALSE,include=TRUE, fig.height = 5, fig.width = 7, fig.align = "center"}
D=rep(1,n)*rgamma(n,25,,5)
H=makeH(NN,D,dx,dy)
ysim=matrix(NA,n,T)
ysim[,1] = y0
for(t in 2:T){
  ysim[,t] = H%*%ysim[,t-1] + r*ysim[,t-1]*(1-ysim[,t-1]/k) #H%*%
}
matplot(t(ysim),type="l",ylab="Cover, %", xlab="Time, years")
```

***  

Non-spatial logistic growth (left), and  spatio-temporal process (right) then looks:  
```{r, echo=FALSE,include=TRUE,message=FALSE, fig.height = 8, fig.width = 11, fig.align = "center"}
library(rasterVis)
library(gridExtra) # also loads grid
D=rep(1,n)*rgamma(n,25,,5)
H=makeH(NN,D,dx,dy)
ysimspat=matrix(NA,n,T)
ysimspat[,1] = y0
ysim=matrix(NA,n,T)
ysim[,1] = y0
for(t in 2:T){
  ysimspat[,t] = H%*%ysimspat[,t-1] + r*ysimspat[,t-1]*(1-ysimspat[,t-1]/k) #H%*%
  ysim[,t] = ysim[,t-1] + r*ysim[,t-1]*(1-ysim[,t-1]/k) 
}
spat = stack()
non = stack()
for(i in 1:12){
  spat=stack(spat,raster(wrap(ysimspat[,i],poly)))
  non=stack(non,raster(wrap(ysim[,i],poly)))
}
p1=levelplot(non,names.attr=paste("Year",c(1:12)), main="Logistic growth")
p2=levelplot(spat,names.attr=paste("Year",c(1:12)), main="IDE based")
grid.arrange(p1,p2, ncol=2,padding=0)
```

***  

Spatio-temporal patterns in the data:
```{r, echo=FALSE,include=TRUE, fig.height = 5, fig.width = 7, fig.align = "center"}
dat = stack()
for(i in 1:12){
  dat=stack(dat,raster(wrap(ymat[,i],poly)))
}
levelplot(dat,names.attr=paste("Year",c(1:12)), main="Data")
```


***  
  
  Both models are mechanistic, but the observed patterns in the adata are different.  

Potential research hypothesis: Our working hypothesis could be that the recovery is driven by diffution, that is propagule dispersal from the edge of the remnant/unburned sagebrush. It is also possible, however, that dispersal is driven, e.g., by long-distance seed dispersal or some unknown spatial stochastic events. This would be expected under "Reid's paradox" where dispersal is faster than expected gradual diffusion. The lack of predictive capacity of the IDE model may be indicative of such processes (or highly variable diffusion rates?) and focusing on spatial patterns of the error would point to areas where demographic processes, compared to biophysical factors, may be less important.

