---
title: "Spatial logistic - stability"
author: "AZ"
date: "21 08 2020"
output:
  html_document: default
  pdf_document: default
---

```{r, include=FALSE}
###### Spatio temporal data simulation
library(raster)
library(rstan)
library(ggplot2)
library(dplyr)
library(viridisLite)
library(reshape2)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

neighborhood <- function(raster){
  nn <- matrix(NA,length(raster[]),4)
  for(i in 1:dim(nn)[1]){
    loc <- adjacent(raster,i)[,2]
    ln <- loc[which((loc+1)==i)]
    rn <- loc[which((loc-1)==i)]
    bn <- loc[which((loc-dim(raster)[2])==i)]
    tn <- loc[which((loc+dim(raster)[2])==i)]
    nn[i,1] <- ifelse(length(ln)>0,ln,0)
    nn[i,2] <- ifelse(length(rn)>0,rn,0)
    nn[i,3] <- ifelse(length(tn)>0,tn,0)
    nn[i,4] <- ifelse(length(bn)>0,bn,0)
  }
  nn
}
neighborhoodMat <- function(raster){
  N <- length(raster[])
  nn <- matrix(0, N, N)
  diag(nn) <- 1:N
  for(i in 1:dim(nn)[1]){
    loc <- adjacent(raster, i)[, 2]
    ln <- loc[which((loc + 1) == i)]
    rn <- loc[which((loc - 1) == i)]
    bn <- loc[which((loc - dim(raster)[2]) == i)]
    tn <- loc[which((loc + dim(raster)[2]) == i)]
    if (length(ln) > 0) { nn[i, i+1] = ln } 
    if (length(rn) > 0) { nn[i, i-1] = rn }
    nn[i,3] <- ifelse(length(tn) > 0, tn, 0)
    nn[i,4] <- ifelse(length(bn) > 0, bn, 0)
  }
  nn
}
wrap <- function(x, poly){
  row = dim(poly)[1]
  col = dim(poly)[2]
  mat=matrix(x, row, col, byrow=TRUE)
  mat
}

### - a function of discretized Fickian diffusion
makeH <- function(NN, Dvec, dx, dy){
  H <- matrix(0, dim(NN)[1], dim(NN)[1])
  diag(H) <- Dvec * (1 / dx^2 + 1 / dy^2)
  for(i in 1:dim(H)[1]){
    H[i,i] = Dvec[i]*(1/dx^2 + 1/dy^2)
    if(sum(NN[i, ] != 0) == 4){
      if(NN[i, 1] > 0){H[i, NN[i, 1]] <- (Dvec[i] - (Dvec[NN[i, 2]] - Dvec[NN[i, 1]]) / 4) / (dx^2)}
      if(NN[i, 2] > 0){H[i, NN[i, 2]] <- (Dvec[i] + (Dvec[NN[i, 2]] - Dvec[NN[i, 1]]) / 4) / (dx^2)}
      if(NN[i, 3] > 0){H[i, NN[i, 3]] <- (Dvec[i] + (Dvec[NN[i, 4]] - Dvec[NN[i, 3]]) / 4) / (dy^2)}
      if(NN[i, 4] > 0){H[i, NN[i, 4]] <- (Dvec[i] - (Dvec[NN[i, 4]] - Dvec[NN[i, 3]]) / 4) / (dy^2)}
    }}
  H
}
makeG <- function(NN, uvec , dx, dy){
  G = matrix(0,dim(NN)[1],dim(NN)[1])
  for(i in 1:dim(H)[1]){
    if(sum(NN[i,]!=0) == 4){
      G[i,i] <- (uvec[NN[i,2]] - 2*uvec[i] + uvec[NN[i,1]])/(2*dx^2) + (uvec[NN[i,4]] - 2*uvec[i] + uvec[NN[i,3]])/(2*dy^2)
      if(NN[i,1]>0){H[i,NN[i,1]] <- -(uvec[NN[i,2]] - uvec[NN[i,1]])/(4*dx^2)}
      if(NN[i,2]>0){H[i,NN[i,2]] <- (uvec[NN[i,2]] - uvec[NN[i,1]])/(4*dx^2)}
      if(NN[i,3]>0){H[i,NN[i,3]] <- -(uvec[NN[i,3]] - uvec[NN[i,3]])/(4*dy^2)}
      if(NN[i,4]>0){H[i,NN[i,4]] <- (uvec[NN[i,4]] - uvec[NN[i,3]])/(4*dy^2)}}
  }
  G
}

### - a function of discretized Plain diffusion with varying D
makeGplain <- function(NN, uvec , dx, dy, dt){
  G <- matrix(0, dim(NN)[1], dim(NN)[1])
  for(i in 1:dim(G)[1]){
    G[i,i] <- -4 * uvec[i]
    if(sum(NN[i, ] != 0) == 4){
      if(NN[i, 1] > 0){G[i, NN[i, 1]] <- uvec[NN[i,1]]}
      if(NN[i, 2] > 0){G[i, NN[i, 2]] <- uvec[NN[i,2]]}
      if(NN[i, 3] > 0){G[i, NN[i, 3]] <- uvec[NN[i,3]]}
      if(NN[i, 4] > 0){G[i, NN[i, 4]] <- uvec[NN[i,4]]}}
  }
  G *(dt / (dx^2 + dy^2))
}
### - a neighborhood matrix 0/1
#makeNNBmatrix <- function()

# --- load data and transform to N x T formal
# years <- c(1985:2017)
# path <- "~/MEGA/landsat_backintime/"
# file <- "BogusCreek_1986_clipped_30_30.tif" #1986
# ras <- stack(paste0(path,file))
# poly <- subset(ras,1:14)
# values(poly)[values(poly) == 101] <- NA
# df1 <- as.data.frame(poly, xy = TRUE, na.rm = TRUE)
# names(df1) <- c('x', 'y', paste0('year', years[1:14]))

# === a setup for deterministic simulation
n <- 15
T <- 9
matsim <- matrix(rnorm(n^2, 1, .1), n, n)
NN0 <- neighborhood(raster(matsim))
bndIdx = which(apply(NN0, 1, function(x) {sum(x == 0) > 0 }))
NN <- NN0
# make diffusion stay within the spatial field
for(i in 1:nrow(NN)){
  for(j in 1:4){
    if(NN[i, j] == 0) {
      NN[i, j] <- i
    }
  }
}
adjMatrix = landscapemetrics::get_adjacencies(raster(matsim), neighbourhood = 4, what = "full")

# mesh parameters
dx <- 30
dy <- 30
dt <- 1

# demographic parameters
r <- .5
k <- 15
D <- rep(50, n^2)

# initiate the simulation
var <- sample(n, n/3)
matsim[var, var] <- k/2 #rnorm(25, 2, .1)
ysim <- matrix(.01, n^2, T) #runif(n^2, 0, k/3)
ysim[, 1] <- as.numeric(matsim)
ysim <- ysim/k

simsage <- function(ysim, r, k, D, dx, dy, dt, NN, spat = TRUE) {
  if(sum((ysim[, 1] > k)) > 0){
    return("error")
  } else{
    for(t in 2:ncol(ysim)){
      #print(t)
      if(spat == TRUE){
        G = makeGplain(NN, ysim[, t-1], dx, dy, dt)
        ysim[, t] = ysim[, t-1] + G %*% D + r * ysim[, t-1] * (1 - ysim[, t-1])
      } else {
        ysim[, t] = ysim[, t-1] + r * ysim[, t-1] * (1 - ysim[, t-1])
      }
    }
  return(ysim)
  }
}

```

### Stability analysis based on CFL criterion  
The reaction-diffusion model is as following:
$$ \frac{\partial u}{\partial t} = 
  \delta(\frac{\partial^2 }{\partial x^2} + \frac{\partial^2 }{\partial y^2})u + ur(1-u)$$
where, $u(t,x,y)$,  $\delta(x,y)$, and r is a constant.  
Then, we can dicretize using a finite difference method with forward difference in time and central in space (assuming $\Delta t = 1$):

$$ 
\begin{aligned}
u_{t+\Delta t,x,y} = &
  u_{t,x,y} + ru_{t,x,y}(1-u_{t,x,y})+ \\
  & \bigg[ \delta_{x,y}(-4u_{t,x,y})+ \\
  & \delta_{x,y}u_{t,x-\Delta x,y}+\delta_{x,y}u_{t,x+\Delta x,y}+ \\
  & \delta_{x,y}u_{t,x,y-\Delta y}+\delta_{x,y}u_{t,x,y+\Delta y} \bigg] \frac{1}{(\Delta x^2 + \Delta y^2)}
\end{aligned}
$$

The stability criterion used is calculated as:
$$ CFL = \frac{\delta_{x,y} * \Delta t}{\Delta x} $$
where $\delta$ is a diffusion rate, $\Delta x = \Delta y$, and the system is stable when $CFL < 1$.

#### Simulations along $\Delta t$, $\delta = 1$:  

```{r, echo=FALSE, message=FALSE}
dt <- c(.1, .5, .75, 1, 10, 100, 150, 200, 365) #seq(.6, 1.4, l = 9)
Dvec <-  rep(1, 9) #seq(1, 500, l = 9) # 
D <- matrix(NA, n^2, T)
par(mfrow = c(3, 3), mar = c(2, 1, 2.5, 1))
for(i in 1:length(Dvec)){
  D[, i] = Dvec[i]
  ysim <- simsage(ysim = ysim, r = .9, k = k, D = D[, i],
                  dx = dx, dy = dy, dt = dt[i], NN = NN,
                  spat = TRUE)
  CFL = round((mean(Dvec[i]) * dt[i]) / dx, 4)
  matplot(t(ysim[-bndIdx,]), type="l" , ylab="Cover, %", xlab="Time, years", col = viridis(n),
          main = paste("dt =", dt[i], ", CFL =", CFL))
}
```
  
#### Simulations along $\delta$, $\Delta t = 1$:  

```{r, echo=FALSE, message=FALSE}
dt <- rep(1, 9) #c(.1, .5, .75, 1, 10, 100, 150, 200, 365) #seq(.6, 1.4, l = 9)
Dvec <-  seq(1, 500, l = 9) # rep(1, 9) #
D <- matrix(NA, n^2, T)
par(mfrow = c(3, 3), mar = c(2, 1, 2.5, 1))
for(i in 1:length(Dvec)){
  D[, i] = Dvec[i]
  ysim <- simsage(ysim = ysim, r = .9, k = k, D = D[, i],
                  dx = dx, dy = dy, dt = dt[i], NN = NN,
                  spat = TRUE)
  CFL = round((mean(Dvec[i]) * dt[i]) / dx, 4)
  matplot(t(ysim[-bndIdx,]), type="l" , ylab="Cover, %", xlab="Time, years", col = viridis(n),
          main = paste("dt =", dt[i], ", CFL =", CFL))
}
```

